---
title: "Assignment 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Ryan Drake"
date: 09/30/2025
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **[Iowa] Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Assignments" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `assignments/assignment_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: assignments/assignment_1/your_file_name.qmd
      text: "Assignment 1: Census Data Exploration"
```
If there is a special character like comma, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)

library(tidycensus)
library(tidyverse)
library(knitr)

# Set your Census API key

census_api_key(Sys.getenv("a55638f442ab081818dbf0ae29a32549253a2ab0"))

# Choose your state for analysis - assign it to a variable called my_state

state_choice <- "IA"

```

**State Selection:** I have chosen **[Iowa]** for this analysis because: [I am interested in rural policy analytics. How is it different than urban. What is often missed in algorithms targeted at rural populations? What are the bias? Data biases? Also I lived there for a few years.]

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r iowa_counties}
# Write your get_acs() code here
# Variables of interest
vars <- c(
  median_household_income = "B19013_001",
  total_population = "B01003_001"
)

# Pull 2022 ACS 5-year data for counties in Iowa
iowa_counties <- get_acs(
  geography = "county",
  variables = c(
    median_household_income = "B19013_001",
    total_population = "B01003_001"
  ),
  state = "IA",
  year = 2022,
  survey = "acs5",
  output = "wide"
) %>%
  mutate(
    county = NAME %>%
      str_remove(", Iowa") %>%
      str_remove(", County")
  ) %>%
  select(
    GEOID,
    county,
    median_household_incomeE,
    median_household_incomeM,   # include MOE column
    total_populationE,
    total_populationM           # include MOE for total population if needed
  )
head(iowa_counties)

# Clean the county names to remove state name and "County" 

# Hint: use mutate() with str_remove()

iowa_counties <- get_acs(
  geography = "county",
  variables = c(
    median_household_income = "B19013_001",
    total_population = "B01003_001"
  ),
  state = "IA",
  year = 2022,
  survey = "acs5",
  output = "wide"
) %>%
  # Create readable county column BEFORE selecting other columns
  mutate(
    county = NAME %>%
      str_remove(", Iowa") %>%
      str_remove(" County")
  ) %>%
  # Now keep only the columns you need, including MOE
  select(
    GEOID,
    county,
    median_household_incomeE,
    median_household_incomeM,
    total_populationE,
    total_populationM
  )

# Display the first few rows

head(iowa_counties)

```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
library(dplyr)

# Calculate MOE percentages
library(stringr)
iowa_reliability <- iowa_counties %>%
  mutate(
    income_moe_pct = (median_household_incomeM / median_household_incomeE) * 100,
    reliability = case_when(
      income_moe_pct < 10 ~ "High",
      income_moe_pct < 20 ~ "Medium",
      TRUE ~ "Low"
    )
  )

reliability_summary <- iowa_reliability %>%
  count(reliability) %>%
  mutate(pct = (n / sum(n)) * 100)


# Create a summary showing count of counties in each reliability category

# Hint: use count() and mutate() to add percentages

reliability_summary
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage

library(knitr)
iowa_reliability <- iowa_counties %>%
  mutate(
    income_moe_pct = (median_household_incomeM / median_household_incomeE) * 100,
    reliability = case_when(
      income_moe_pct < 10 ~ "High",
      income_moe_pct < 20 ~ "Moderate",
      TRUE ~ "Low"
    )
  )
top5_moe <- iowa_reliability %>%
  arrange(desc(income_moe_pct)) %>%
  slice_head(n = 5) %>%
  select(
    County = county,
    `Median Household Income` = median_household_incomeE,
    `Margin of Error` = median_household_incomeM,
    `MOE %` = income_moe_pct,
    Reliability = reliability
  )


# Format as table with kable() - include appropriate column names and caption
kable(
  top5_moe,
  caption = "Top 5 Iowa Counties by Median Household Income MOE Percentage",
  digits = 1
)


```

**Data Quality Commentary:**

[Write 2-3 sentences explaining what these results mean for algorithmic decision-making. Consider: Which counties might be poorly served by algorithms that rely on this income data? What factors might contribute to higher uncertainty?]

[The top five reliability in such a rural state only have Medium qualifiers. None are Low. The data is not that reliable and could prove difficult for algorithmic decision-making in rural areas. Scale is perhaps a big thing because of course population is lower so decreases reliability but it was an interesting test to see what MoE you would get in these areas. Also surprisingly it is not Des Moines county or Ames county which are much larger.]

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties

selected_counties <- iowa_reliability %>%
  filter(county %in% c("Adams", "Monroe")) %>%
  select(
    County = county,
    `Median Household Income` = median_household_incomeE,
    `MOE %` = income_moe_pct,
    Reliability = reliability
  )

# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category

selected_counties
```

**Comment on the output:** [Its just a smaller tibble.]

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names

race_vars <- c(
  white = "B03002_003",
  black = "B03002_004",
  hispanic = "B03002_012",
  total = "B03002_001"
)

# Use get_acs() to retrieve tract-level data
county_codes <- str_sub(iowa_counties$GEOID, 3, 5)

# Hint: You may need to specify county codes in the county parameter

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations
iowa_tracts <- get_acs(
  geography = "tract",
  variables = race_vars,
  state = "IA",
  county = county_codes,
  year = 2022,
  survey = "acs5",
  output = "wide"
) %>%
  mutate(
    pct_white = (whiteE / totalE) * 100,
    pct_black = (blackE / totalE) * 100,
    pct_hispanic = (hispanicE / totalE) * 100,
    # Extract county and tract names separately
    county_name = str_extract(NAME, "[^,]+"),
    tract_name = str_extract(NAME, "Census Tract \\d+\\.?\\d*")
  ) %>%
  select(GEOID, tract_name, county_name,
         totalE, whiteE, whiteM, blackE, blackM, hispanicE, hispanicM,
         pct_white, pct_black, pct_hispanic)


# Add readable tract and county name columns using str_extract() or similar

head(iowa_tracts)

```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
top_hispanic_tract <- iowa_tracts %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) %>%
  select(tract_name, county_name, pct_hispanic, totalE, hispanicE)

top_hispanic_tract


# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group

county_summary <- iowa_tracts %>%
  group_by(county_name) %>%
  summarize(
    num_tracts = n(),
    avg_pct_white = mean(pct_white, na.rm = TRUE),
    avg_pct_black = mean(pct_black, na.rm = TRUE),
    avg_pct_hispanic = mean(pct_hispanic, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_pct_hispanic)) 

# Create a nicely formatted table of your results using kable()
kable(
  county_summary,
  caption = "Iowa County-Level Average Racial/Ethnic Percentages and Tract Counts",
  digits = 1
)

```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
iowa_tracts <- iowa_tracts %>%
  mutate(
    moe_pct_white = (whiteM / whiteE) * 100,
    moe_pct_black = (blackM / blackE) * 100,
    moe_pct_hispanic = (hispanicM / hispanicE) * 100
  )

# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement
iowa_tracts_qc <- iowa_tracts %>%
  mutate(
    moe_pct_white = (whiteM / whiteE) * 100,
    moe_pct_black = (blackM / blackE) * 100,
    moe_pct_hispanic = (hispanicM / hispanicE) * 100,
    high_moe_flag = ifelse(
      moe_pct_white > 20 | moe_pct_black > 20 | moe_pct_hispanic > 20,
      TRUE,
      FALSE
    )
  )

# Create summary statistics showing how many tracts have data quality issues
moe_summary <- iowa_tracts_qc %>%
  summarize(
    total_tracts = n(),
    tracts_high_moe = sum(high_moe_flag),
    pct_high_moe = (tracts_high_moe / total_tracts) * 100
  )

moe_summary

```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
iowa_tracts <- iowa_tracts %>%
  mutate(
    high_moe_flag = ifelse(
      moe_pct_white > 20 | moe_pct_black > 20 | moe_pct_hispanic > 20,
      TRUE,
      FALSE
    )
  )

# Calculate average characteristics for each group:
# - population size, demographic percentages
# Use group_by() and summarize() to create this comparison
moe_comparison <- iowa_tracts %>%
  group_by(high_moe_flag) %>%
  summarize(
    num_tracts = n(),
    avg_population = mean(totalE, na.rm = TRUE),
    avg_pct_white = mean(pct_white, na.rm = TRUE),
    avg_pct_black = mean(pct_black, na.rm = TRUE),
    avg_pct_hispanic = mean(pct_hispanic, na.rm = TRUE)
  )

# Create a professional table showing the patterns

kable(
  moe_comparison,
  caption = "Comparison of Census Tracts by High MOE Flag",
  digits = 1
)
```

**Pattern Analysis:** [Describe any patterns you observe. Do certain types of communities have less reliable data? What might explain this?]

[It makes sense to me that areas with smaller total populations will have a higher MoE. Racially also if the population is smaller, like in Iowa - Black populations are very small, they would have higher MoE as well than their White communities. ]


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**

[1. Rural areas still contain many High MoE at the county level. When you transfer that to neighborhoods or groups of people, the MoE just increases (gets lower) because the sample size gets even smaller. This could pose potential issues in algorithmic analysis for policy recommendations and you really need to monitor the outcomes.
2. Smaller communities, like for instance Iowa has a very low population of Black people, would have extremely high MoE and would face greater risk of algorithmic bias.
3. The sample size. We also talked about human choice in all the steps like how you clean data, what you choose to classify (like MoE as high or medium what percent makes this decision?), to variables you put into the model.
4. There definitely needs to be checks as there are a lot of blind spots, especially it seems in rural areas where data quality and quantity are not as good as urban areas? One needs guardrails and other modes of analysis on top of algorithmic analysis.]

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category
county_reliability <- iowa_reliability %>%
  mutate(
    income_moe_pct = (median_household_incomeM / median_household_incomeE) * 100,
    reliability = case_when(
      income_moe_pct < 10 ~ "High",
      income_moe_pct < 20 ~ "Moderate",
      TRUE ~ "Low"
    ),

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"
    algorithm_recommendation = case_when(
      reliability == "High" ~ "Safe for algorithmic decisions",
      reliability == "Moderate" ~ "Use with caution - monitor outcomes",
      reliability == "Low" ~ "Requires manual review or additional data"
    )
  ) %>%
  select(
    County = county,
    `Median Household Income` = median_household_incomeE,
    `MOE %` = income_moe_pct,
    Reliability = reliability,
    `Algorithm Recommendation` = algorithm_recommendation
  )

# Format as a professional table with kable()
kable(
  county_reliability,
  caption = "Iowa County Median Income Reliability and Algorithm Recommendations",
  digits = 1
)

```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** [Most counties in Iowa are save for immediate algorithmic implementation because our model describes High MoE as less than 10%. Most fit this category.]

2. **Counties requiring additional oversight:** [There are just a few that are listed as Moderate and would require additional oversight including Adams, Audubon, Cass, Clay, Howard, Ida, Louisa, Lucas, Monroe, O'Brien, Ringgold, Shelby, Wayne, and Winnebago]

3. **Counties needing alternative approaches:** [There were actually no counties needed alternative approaches. This was a bit surprising as I thought most has potential to fall into this category.]

## Questions for Further Investigation

[Why were there no Low scoring counties? Is that because our parameters for scoring these categories are too low? This was my own human error? Are county level analysis usually safe for algorithmic implementation anyway?]

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on [date]

**Reproducibility:** 
- All analysis conducted in R version [your version]
- Census API key required for replication
- Complete code and documentation available at: [your portfolio URL]

**Methodology Notes:**
[I may have way over simplified the parameters of High, Medium, and Low MoE. I will need to check this in class.]

**Limitations:**
[I thought the geographic (rural) scope would be interesting to do but I don't think it gave any limitations as the sample size for the county level and neighborhood level seemed usable. The data is simply from ACS.]

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable 
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/assignments/assignment_1/your_file_name.html`