---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Ryan Drake"
date: 10/14/2025
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**
- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data:
- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
# Load required packages
library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)
library(ggplot2)
# Load spatial data

census_api_key("42bf8a20a3df1def380f330cf7edad0dd5842ce6")

pa_counties <- st_read(here("Assignments/assignment_02/data/Pennsylvania_County_Boundaries.shp"))
districts <- st_read(here("Assignments/assignment_02/data/districts.geojson"))
hospitals <- st_read(here("Assignments/assignment_02/data/hospitals.geojson"))
metro_areas <- core_based_statistical_areas(cb = TRUE)

# Check that all data loaded correctly
hospitals <- hospitals %>%
  st_transform(st_crs(pa_counties))

library(tigris)
options(tigris_use_cache = TRUE)
census_tracts <- tracts(state = "PA", cb = TRUE, year = 2024)

#make crs match I'll review this later!
census_tracts <- census_tracts %>%
  st_transform(st_crs(pa_counties))

st_crs(pa_counties)
st_crs(hospitals)
st_crs(census_tracts)
st_crs(districts)
st_crs(metro_areas)

```

**Questions to answer:**
- How many hospitals are in your dataset?
    - There are 223 observations of hospitals in the dataset. 
- How many census tracts?
    - There are 3445 census tracts observed.
- What coordinate reference system is each dataset in?
    - They are all in WGS 84. Except Metro which is is NAD83. 

---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**
- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
# Get demographic data from ACS
# Define the ACS year and dataset
year <- 2021
survey <- "acs5"

# Get tract-level demographic data for Pennsylvania
pa_demo <- get_acs(
  geography = "tract",
  state = "PA",
  year = year,
  survey = survey,
  variables = c(
    total_pop = "B01003_001",         # Total population
    median_income = "B19013_001",     # Median household income (dollars)
    age_65_66 = "B01001_020",         # Male 65-66
    age_67_69 = "B01001_021",         # Male 67-69
    age_70_74 = "B01001_022",         # Male 70-74
    age_75_79 = "B01001_023",         # Male 75-79
    age_80_84 = "B01001_024",         # Male 80-84
    age_85_up = "B01001_025",         # Male 85+
    f_age_65_66 = "B01001_044",       # Female 65-66
    f_age_67_69 = "B01001_045",       # Female 67-69
    f_age_70_74 = "B01001_046",       # Female 70-74
    f_age_75_79 = "B01001_047",       # Female 75-79
    f_age_80_84 = "B01001_048",       # Female 80-84
    f_age_85_up = "B01001_049"        # Female 85+
  ),
  geometry = FALSE  # Set TRUE later if you want spatial data
)
pa_demo_summary <- pa_demo %>%
  select(GEOID, variable, estimate) %>%
  tidyr::pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    pop_65_over = age_65_66 + age_67_69 + age_70_74 +
                  age_75_79 + age_80_84 + age_85_up +
                  f_age_65_66 + f_age_67_69 + f_age_70_74 +
                  f_age_75_79 + f_age_80_84 + f_age_85_up
  ) %>%
  select(GEOID, total_pop, median_income, pop_65_over)

head(pa_demo_summary)


# Join to tract boundaries
pa_tracts <- tracts(state = "PA", cb = TRUE, year = 2024)
pa_demo_joined <- pa_tracts %>%
  left_join(pa_demo_summary, by = "GEOID")

pa_demo_joined
plot(pa_demo_joined["median_income"])

```

**Questions to answer:**
- What year of ACS data are you using?
    - I tried to use the most recent data so I used 2024. I tried 2025 first but kept getting a Census Bureau error since that data isn't out yet obviously.I was just curious what it would do if I tried. 
    
- How many tracts have missing income data?

# Count tracts with missing median household income

missing_income_count <- pa_demo_joined %>%
  summarize(
    n_missing_income = sum(is.na(median_income)),
    total_tracts = n(),
    percent_missing = 100 * n_missing_income / total_tracts
  )

missing_income_count


A total of 65 tracts have missing income. 

- What is the median income across all PA census tracts?

# median of tracts median household incomes
median_income_pa <- pa_demo_joined %>%
  summarize(
    state_median_income = median(median_income, na.rm = TRUE)
  )

median_income_pa

The median income for all tracts is $65195.50. 

---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income (choose an appropriate threshold)
2. Significant elderly population (choose an appropriate threshold)

**Your Task:**
```{r}
# Filter for vulnerable tracts based on your criteria
pa_demo_joined <- pa_demo_joined %>%
  mutate(
    pct_65_over = 100 * pop_65_over / total_pop
  )
income_threshold <- 40000
elderly_threshold <- 20

vulnerable_tracts <- pa_demo_joined %>%
  filter(
    median_income <= income_threshold,
    pct_65_over >= elderly_threshold
  )

summary(vulnerable_tracts$median_income)
summary(vulnerable_tracts$pct_65_over)
nrow(vulnerable_tracts)

```

**Questions to answer:**
- What income threshold did you choose and why? 

    I didn't want to choose the poverty line because individuals (or families) who make just over can still be significantly impacted by income while not being eligible for benifits of programs. Though this is all of PA and not everywhere is as expensive as Philadelphia or Pittsburg but still thought it a fine threshhold. 
    
- What elderly population threshold did you choose and why? 

  This one I had to google a bit to find a suitable threshold. I found that setting a threshold of 20, where 20% of the population is aged 65+ is considered "significant elderly population."
  
- How many tracts meet your vulnerability criteria?

95 tracts in total. 

- What percentage of PA census tracts are considered vulnerable by your definition?

2.75% 

---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}
# Transform to appropriate projected CRS

#NAD83
pa_tracts_proj <- st_transform(pa_demo_joined, 26918)
vulnerable_centroids <- vulnerable_tracts %>%
  st_transform(26918) %>%
  st_centroid()

# Calculate distance from each tract centroid to nearest hospital

hospitals_proj <- st_transform(hospitals, 26918)

dist_matrix <- st_distance(vulnerable_centroids, hospitals_proj)

vulnerable_centroids <- vulnerable_centroids %>%
  mutate(
    dist_to_nearest_hospital_m = apply(dist_matrix, 1, min),
    dist_to_nearest_hospital_miles = dist_to_nearest_hospital_m * 0.000621371
  )

summary(vulnerable_centroids$dist_to_nearest_hospital_miles)
head(vulnerable_centroids[, c("GEOID", "dist_to_nearest_hospital_miles")])

```

**Requirements:**
- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles
- Explain why you chose your projection

**Questions to answer:**
- What is the average distance to the nearest hospital for vulnerable tracts?

The average distance to the nearest hospital is 2.61 miles. 

- What is the maximum distance?

The maximum distance is 18.68 miles. 

- How many vulnerable tracts are more than 15 miles from the nearest hospital?

vulnerable_far <- vulnerable_centroids %>%
  filter(dist_to_nearest_hospital_miles > 15)
  
summary_far <- vulnerable_far %>%
  summarize(
    n_far = n(),
    total_vulnerable = nrow(vulnerable_centroids),
    percent_far = 100 * n_far / total_vulnerable
  )

summary_far

There is only one tract that is more than 15 miles away from the nearest hospital. 
---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable

vulnerable_centroids <- vulnerable_centroids %>%
  mutate(
    underserved = dist_to_nearest_hospital_miles > 15
  )

underserved_summary <- vulnerable_centroids %>%
  summarize(
    n_underserved = sum(underserved, na.rm = TRUE),
    total_vulnerable = n(),
    percent_underserved = 100 * n_underserved / total_vulnerable
  )

underserved_summary

```

**Questions to answer:**
- How many tracts are underserved?

  There is still jsut one underserved tract. 
  
- What percentage of vulnerable tracts are underserved?

  1%

- Does this surprise you? Why or why not?

Yes, I thought there would be more in more rural areas of PA. 

---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
# Spatial join tracts to counties
vulnerable_centroids <- st_transform(vulnerable_centroids, st_crs(pa_counties))
pa_counties <- st_transform(pa_counties, st_crs(vulnerable_centroids))

vulnerable_with_county <- st_join(vulnerable_centroids, pa_counties, join = st_within)

# Aggregate statistics by county

county_stats <- vulnerable_with_county %>%
  st_drop_geometry() %>%
  group_by(county_name = NAMELSADCO) %>% 
  summarize(
    n_vulnerable_tracts = n(),
    n_underserved_tracts = sum(underserved, na.rm = TRUE),
    pct_underserved = 100 * n_underserved_tracts / n_vulnerable_tracts,
    avg_dist_miles = mean(dist_to_nearest_hospital_miles, na.rm = TRUE),
    total_vulnerable_pop = sum(total_pop, na.rm = TRUE)
  ) %>%
  arrange(desc(n_underserved_tracts))
head(county_stats)

```

**Required county-level statistics:**
- Number of vulnerable tracts
- Number of underserved tracts  
- Percentage of vulnerable tracts that are underserved
- Average distance to nearest hospital for vulnerable tracts
- Total vulnerable population

**Questions to answer:**
- Which 5 counties have the highest percentage of underserved vulnerable tracts?

     Cameron County, 100%
    
- Which counties have the most vulnerable people living far from hospitals?
  
    Allegheny, Cambria, Beaver, Vlair, Berks, and Cameron

- Are there any patterns in where underserved counties are located?

  Mostly they are rural counties except for Allegheny. 

---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
# Create and format priority counties table

county_stats <- county_stats %>%
  mutate(
    pct_vulnerable = 100 * n_vulnerable_tracts / sum(n_vulnerable_tracts)
  )

#Put Cameron on the top regardless of vulnerable rank
cameron <- county_stats %>%
  filter(county_name == "Cameron") %>%
  mutate(county_name = paste0("**", county_name, "**"))  
others <- county_stats %>%
  filter(county_name != "Cameron") %>%
  arrange(desc(pct_vulnerable))
priority_counties <- bind_rows(cameron, others)

#Select top 10 counties by % vulnerable tracts
priority_counties <- priority_counties %>%
  select(
    county_name,
    n_vulnerable_tracts,
    n_underserved_tracts,
    pct_vulnerable,
    avg_dist_miles,
    total_vulnerable_pop
  ) %>%
  mutate(
    pct_vulnerable = round(pct_vulnerable, 1),
    avg_dist_miles = round(avg_dist_miles, 1),
    total_vulnerable_pop = prettyNum(total_vulnerable_pop, big.mark = ",")
  )

#Display the table
library(knitr)
priority_counties_top10 <- priority_counties %>% slice_head(n = 10)
kable(priority_counties_top10,
      caption = "Top 10 Priority Counties for Healthcare Investment in Pennsylvania",
      col.names = c("County",
                    "Vulnerable Tracts",
                    "Underserved Tracts",
                    "% Vulnerable",
                    "Avg Distance to Hospital (miles)",
                    "Total Vulnerable Population"),
      align = c("l","r","r","r","r","r"),
      format = "html")
```

**Requirements:**
- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
# Create county-level access map
pa_counties <- pa_counties %>%
  rename(county_name = COUNTY_NAM)

pa_counties_map <- pa_counties %>%
  left_join(county_stats %>% 
              select(county_name, pct_underserved),
            by = c("county_name" = "county_name"))

hospitals_proj <- st_transform(hospitals, st_crs(pa_counties_map))

ggplot() +
  geom_sf(data = pa_counties_map, aes(fill = pct_underserved), color = "gray60", size = 0.3) +
  geom_sf(data = hospitals_proj, color = "red", size = 2, shape = 21, fill = "white") +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    na.value = "white",
    labels = scales::percent_format(scale = 1),
    name = "% Vulnerable Tracts\nUnderserved"
  ) +
  theme_void() +
  labs(
    title = "Healthcare Access Challenges in Pennsylvania",
    subtitle = "County-level % of vulnerable tracts that are underserved",
    caption = "Red points indicate hospital locations"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )
```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved
- Include hospital locations as points
- Use an appropriate color scheme
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map
vulnerable_tracts_proj <- st_transform(vulnerable_tracts, st_crs(pa_counties))
pa_counties_proj <- st_transform(pa_counties, st_crs(vulnerable_tracts))
hospitals_proj <- st_transform(hospitals, st_crs(vulnerable_tracts))
vulnerable_tracts_proj <- st_transform(vulnerable_tracts, st_crs(pa_counties))

underserved_summary <- underserved_summary %>%
  rename(underserved = n_underserved)

ulnerable_tracts_proj <- vulnerable_tracts_proj %>%
  left_join(
    vulnerable_centroids %>% st_drop_geometry() %>% select(GEOID, underserved),
    by = "GEOID"
  )

underserved_tracts <- ulnerable_tracts_proj %>%
  filter(underserved == TRUE)

ggplot() +
  # County boundaries (subtle gray)
  geom_sf(data = pa_counties_proj, fill = NA, color = "gray50", size = 0.4) +
  
  # All vulnerable tracts (light fill, low alpha)
  geom_sf(data = ulnerable_tracts_proj, fill = "lightblue", color = NA, alpha = 0.3) +
  
  # Underserved vulnerable tracts (stand out)
  geom_sf(data = underserved_tracts, fill = "red", color = NA, alpha = 0.7) +
  
  # Hospital locations
  geom_sf(data = hospitals_proj, shape = 21, color = "black", fill = "white", size = 2) +
  
  # Titles and labels
  labs(
    title = "Underserved Vulnerable Census Tracts in Pennsylvania",
    subtitle = "Red tracts are vulnerable and more than 15 miles from the nearest hospital",
    caption = "County boundaries shown in gray; hospitals shown as white circles with black border"
  ) +
  
  # Clean map
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 10)
  )

```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color
- Include county boundaries for context
- Show hospital locations
- Use appropriate visual hierarchy (what should stand out?)
- Include informative title and subtitle

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization
vulnerable_distances <- vulnerable_centroids %>%
  st_drop_geometry() %>%
  filter(!is.na(dist_to_nearest_hospital_miles))

hist_plot <- ggplot(vulnerable_distances, aes(x = dist_to_nearest_hospital_miles)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "gray40", alpha = 0.8) +
  geom_vline(xintercept = 15, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Distances to Nearest Hospital for Vulnerable Populations",
    subtitle = "Red dashed line indicates 15-mile underserved threshold",
    x = "Distance to Nearest Hospital (miles)",
    y = "Number of Vulnerable Tracts",
    caption = "Most vulnerable tracts are within 15 miles, but several are farther, indicating potential access gaps."
  ) +
  theme_minimal(base_size = 12) +
  scale_x_continuous(breaks = seq(0, max(vulnerable_distances$dist_to_nearest_hospital_miles), by = 5)) +
  scale_y_continuous(labels = comma)
top_counties <- vulnerable_distances %>%
  group_by(NAMELSADCO) %>%
  summarise(total_vulnerable_pop = sum(total_pop, na.rm = TRUE)) %>%
  arrange(desc(total_vulnerable_pop)) %>%
  slice_head(n = 10) %>%
  pull(NAMELSADCO)

box_plot <- vulnerable_distances %>%
  filter(NAMELSADCO %in% top_counties) %>%
  ggplot(aes(x = reorder(NAMELSADCO, dist_to_nearest_hospital_miles, median), 
             y = dist_to_nearest_hospital_miles)) +
  geom_boxplot(fill = "lightblue", color = "gray40", alpha = 0.8) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red", size = 0.8) +
  coord_flip() +
  labs(
    title = "Distance to Nearest Hospital by Top 10 Counties (Vulnerable Population)",
    x = "County",
    y = "Distance to Nearest Hospital (miles)",
    caption = "Red dashed line marks the 15-mile threshold for underserved tracts."
  ) +
  theme_minimal(base_size = 12)
  

hist_plot
# To display box plot instead
box_plot

```

**Suggested chart types:**
- Histogram or density plot of distances
- Box plot comparing distances across regions
- Bar chart of underserved tracts by county
- Scatter plot of distance vs. vulnerable population size

**Requirements:**
- Clear axes labels with units
- Appropriate title
- Professional formatting
- Brief interpretation (1-2 sentences as a caption or in text)

---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own research question using OpenDataPhilly data (https://opendataphilly.org/datasets/). 

**Note these are just loose suggestions to spark ideas - follow or make your own as the data permits and as your ideas evolve. This analysis should include bringing in your own dataset, ensuring the projection/CRS of your layers align and are appropriate for the analysis (not lat/long or geodetic coordinate systems). The analysis portion should include some combination of spatial and attribute operations to answer a relatively straightforward question**

---

#### Education & Youth Services

**Option A: Educational Desert Analysis**
- **Data:** Schools, Libraries, Recreation Centers, Census tracts (child population)
- **Question:** "Which neighborhoods lack adequate educational infrastructure for children?"
- **Operations:** Buffer schools/libraries (0.5 mile walking distance), identify coverage gaps, overlay with child population density
- **Policy relevance:** School district planning, library placement, after-school program siting

**Option B: School Safety Zones**
- **Data:** Schools, Crime Incidents, Bike Network
- **Question:** "Are school zones safe for walking/biking, or are they crime hotspots?"
- **Operations:** Buffer schools (1000ft safety zone), spatial join with crime incidents, assess bike infrastructure coverage
- **Policy relevance:** Safe Routes to School programs, crossing guard placement

---

#### Environmental Justice

**Option C: Green Space Equity** 
- **Data:** Parks, Street Trees, Census tracts (race/income demographics)
- **Question:** "Do low-income and minority neighborhoods have equitable access to green space?"
- **Operations:** Buffer parks (10-minute walk = 0.5 mile), calculate tree canopy or park acreage per capita, compare by demographics
- **Policy relevance:** Climate resilience, environmental justice, urban forestry investment
---

#### Public Safety & Justice

**Option D: Crime & Community Resources**
- **Data:** Crime Incidents, Recreation Centers, Libraries, Street Lights
- **Question:** "Are high-crime areas underserved by community resources?"
- **Operations:** Aggregate crime counts to census tracts or neighborhoods, count community resources per area, spatial correlation analysis
- **Policy relevance:** Community investment, violence prevention strategies
---

#### Infrastructure & Services

**Option E: Polling Place Accessibility**
- **Data:** Polling Places, SEPTA stops, Census tracts (elderly population, disability rates)
- **Question:** "Are polling places accessible for elderly and disabled voters?"
- **Operations:** Buffer polling places and transit stops, identify vulnerable populations, find areas lacking access
- **Policy relevance:** Voting rights, election infrastructure, ADA compliance

---

#### Health & Wellness

**Option F: Recreation & Population Health**
- **Data:** Recreation Centers, Playgrounds, Parks, Census tracts (demographics)
- **Question:** "Is lack of recreation access associated with vulnerable populations?"
- **Operations:** Calculate recreation facilities per capita by neighborhood, buffer facilities for walking access, overlay with demographic indicators
- **Policy relevance:** Public health investment, recreation programming, obesity prevention

---

#### Emergency Services

**Option G: EMS Response Coverage**
- **Data:** Fire Stations, EMS stations, Population density, High-rise buildings
- **Question:** "Are population-dense areas adequately covered by emergency services?"
- **Operations:** Create service area buffers (5-minute drive = ~2 miles), assess population coverage, identify gaps in high-density areas
- **Policy relevance:** Emergency preparedness, station siting decisions

---

#### Arts & Culture

**Option H: Cultural Asset Distribution**
- **Data:** Public Art, Museums, Historic sites/markers, Neighborhoods
- **Question:** "Do all neighborhoods have equitable access to cultural amenities?"
- **Operations:** Count cultural assets per neighborhood, normalize by population, compare distribution across demographic groups
- **Policy relevance:** Cultural equity, tourism, quality of life, neighborhood identity

---

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/
- Most datasets available as GeoJSON, Shapefile, or CSV with coordinates
- Always check the Metadata for a data dictionary of the fields.

**Additional Sources:**
- **Pennsylvania Open Data:** https://data.pa.gov/
- **Census Bureau (via tidycensus):** Demographics, economic indicators, commute patterns
- **TIGER/Line (via tigris):** Geographic boundaries

### Recommended Starting Points

**If you're feeling confident:** Choose an advanced challenge with multiple data layers. 
**If you are a beginner, choose something more manageable that helps you understand the basics**


**If you have a different idea:** Propose your own question! Just make sure:
- You can access the spatial data
- You can perform at least 2 spatial operations

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
# Load your additional dataset

state_parks <- st_read(here("Assignments/assignment_02/data/DCNR_StateParks202503.geojson"))
state_parks <- st_transform(state_parks, crs = 2272)

st_crs(state_parks)

```

**Questions to answer:**
- What dataset did you choose and why?
  I chose the DCNR State Parks data set to see state park acess and vizualize distribution per captia.
  
- What is the data source and date?
  DCNR and 2025. 

- How many features does it contain?
It contains 12 features. 

- What CRS is it in? Did you need to transform it?

It is in NAD83, I think this is fine for PA. 

---

2. **Pose a research question**

What is the state park acreage per capita by county in Pennsylvania? Which counties have the most state park acerages? Where are there Park and Ride locations in the state which are usually a way for underserved populations to access state parks? 

---

3. **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
# Your spatial analysis
state_parks <- st_read(here("Assignments/assignment_02/data/DCNR_StateParks202503.geojson"))
state_parks <- st_transform(state_parks, crs = 2272)

#get acreage 
state_parks <- state_parks %>%
  mutate(area_sqft = st_area(geometry),
         area_acres = as.numeric(area_sqft) / 43560)

pa_counties <- st_transform(pa_counties, st_crs(state_parks))

parks_by_county <- st_intersection(state_parks, pa_counties) %>%
  st_drop_geometry() %>%
  group_by(county_name) %>%
  summarise(total_park_acres = sum(area_acres, na.rm = TRUE))

#total poulation by county
pa_population <- get_acs(
  geography = "county",
  variables = "B01003_001",  # total population
  state = "PA",
  year = 2023,               # or most recent available
  geometry = FALSE           # set TRUE if you want sf object
) %>%
  select(county_name = NAME, total_population = estimate)

pa_population <- pa_population %>%
  mutate(
    county_name = str_replace(county_name, ", Pennsylvania", ""),  # remove state
    county_name = str_replace(county_name, " County", ""),         # remove "County"
    county_name = str_trim(county_name)                             # remove extra spaces
  )
parks_by_county <- parks_by_county %>%
  mutate(county_name = str_to_lower(county_name))

pa_population <- pa_population %>%
  mutate(county_name = str_to_lower(county_name))

pa_counties <- pa_counties %>%
  mutate(county_name = str_to_lower(county_name))


parks_per_capita <- parks_by_county %>%
  left_join(pa_population %>% select(county_name, total_population), by = "county_name") %>%
  mutate(acres_per_capita = total_park_acres / total_population)

parks_per_capita_sf <- pa_counties %>%
  left_join(parks_per_capita, by = "county_name")

#map parks per capita
pa_counties_map_two <- parks_per_capita_sf

library(scales)
#map
ggplot(pa_counties_map_two) +
  geom_sf(aes(fill = acres_per_capita), color = "grey", size = 0.3) +
  # Gradient coloring
  scale_fill_viridis_c(
    option = "plasma",      # vibrant color palette
    direction = -1,          # high values = darker/brighter
    na.value = "white",
    labels = comma,
    name = "Park Acres per Capita"
  ) +
  
  # Titles and caption
  labs(
    title = "State Park Acres per Capita by County in Pennsylvania",
    subtitle = "Higher values indicate more park land per person",
    caption = "Data: PA State Parks & ACS County Population"
  ) +
  
  # Clean theme
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

```

```{r}
park_ride <- st_read(here("Assignments/assignment_02/data/Park_and_Ride_Locations_Current_Transportation_20251014.geojson"))
park_ride <- st_transform(park_ride, crs = 2272)

park_ride_sf <- st_as_sf(park_ride, coords = c("Longitude", "Latitude"), crs = 4326)

park_ride_sf <- park_ride_sf %>%
  rename(code = dot_county_code)

parks_per_capita_sf <- parks_per_capita_sf %>%
  rename(code = COUNTY_N_1)

joined_sf <- st_join(park_ride_sf, parks_per_capita_sf, by = "code")

pa_counties_map_three <- joined_sf

library(scales)
#map
ggplot(pa_counties_map_three) +
  geom_sf(data = pa_counties, fill = "gray90", color = "white") +
  geom_sf(data = joined_sf, color = "blue", size = 2) +
  theme_minimal() +
  
  geom_sf(aes(fill = acres_per_capita), color = "grey", size = 0.3) +
  # Gradient coloring
  scale_fill_viridis_c(
    option = "plasma",      # vibrant color palette
    direction = -1,          # high values = darker/brighter
    na.value = "white",
    labels = comma,
    name = "Park Acres per Capita"
  ) +
  
  geom_sf(color = "blue", size = 2) +
  labs(
    title = "Park-and-Ride Locations in Pennsylvania",
    subtitle = "Pennsylvania Open Data Portal",
    caption = "Data: PA Open Data Portal"
  ) +
  theme_minimal()

  # Titles and caption
  labs(
    title = "Park and Ride Locations and State Park Acres per Capita by County in Pennsylvania",
    subtitle = "Higher values indicate more park land per person",
    caption = "Data: PA State Parks & ACS County Population"
  ) +

  # Clean theme
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )


```


**Analysis requirements:**
- Clear code comments explaining each step
- Appropriate CRS transformations
- Summary statistics or counts
- At least one map showing your findings
- Brief interpretation of results (3-5 sentences)

**Your interpretation:**

[I was curious to see the distribution of state park land per capita in Pennsylvania. Surprisingly, it is pretty evenly distributed. With just a few counties missing data.The park and ride locations however are not as evenly distributed. This would be a policy recommendation to continue to look into distribution for access sake. (I could not get them on the same map!) ]

## Finally - A few comments about your incorporation of feedback!
Take a few moments to clean up your markdown document and then write a line or two or three about how you may have incorporated feedback that you recieved after your first assignment. 


I am still confused if we are supposed to keep much of the instructions for the website version? It is a really long document and if this is for a portfolio it would need to be heartily cleaned up and presentable? How much of this do we do? 

---

## Submission Requirements

**What to submit:**

1. **Rendered HTML document posted to your course portfolio** with all code, outputs, maps, and text
   - Use `embed-resources: true` in YAML so it's a single file
   - All code should run without errors
   - All maps and charts should display correctly

**File naming:** `LastName_FirstName_Assignment2.html` and `LastName_FirstName_Assignment2.qmd`

---

